---
title: "10 Insights from Alabama School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{10 Insights from Alabama School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  cache = TRUE
)

# Detect if we're in CI/R CMD check - skip network calls but still show plots
in_ci <- Sys.getenv("CI") == "true" || Sys.getenv("_R_CHECK_FORCE_SUGGESTS_") != ""
skip_network <- in_ci
```

```{r load-packages}
library(alschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 14))
```

This vignette explores Alabama's public school enrollment data, surfacing key trends and demographic patterns across 10 years of data (2015-2024).

---

## 1. Alabama's enrollment is holding steady

Unlike many states seeing sharp pandemic-driven declines, Alabama's public school enrollment has remained relatively stable around 730,000 students.

```{r statewide-trend}
if (skip_network) {
  # Use example data during CI
  enr <- alschooldata:::create_example_data()
} else {
  enr <- fetch_enr_multi(2015:2025, use_cache = TRUE)
}

state_totals <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(change = n_students - lag(n_students),
         pct_change = round(change / lag(n_students) * 100, 2))

state_totals
```

```{r statewide-chart}
ggplot(state_totals, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#9B1B30") +
  geom_point(size = 3, color = "#9B1B30") +
  scale_y_continuous(labels = scales::comma, limits = c(700000, 750000)) +
  scale_x_continuous(breaks = 2015:2025) +
  labs(
    title = "Alabama Public School Enrollment (2015-2025)",
    subtitle = "Statewide enrollment has remained remarkably stable",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

---

## 2. COVID hit elementary hardest

The pandemic's enrollment impact was felt most sharply in elementary grades, especially kindergarten, which saw significant drops in 2021 as families delayed school entry.

```{r covid-grades}
covid_grades <- enr |>
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "01", "06", "09"),
         end_year %in% 2019:2023) |>
  select(end_year, grade_level, n_students) |>
  pivot_wider(names_from = end_year, values_from = n_students) |>
  mutate(change_2019_2021 = `2021` - `2019`,
         pct_drop = round(change_2019_2021 / `2019` * 100, 1))

covid_grades
```

```{r covid-chart}
enr |>
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "01", "06", "09"),
         end_year %in% 2019:2023) |>
  mutate(grade_level = factor(grade_level,
                              levels = c("K", "01", "06", "09"),
                              labels = c("Kindergarten", "1st Grade", "6th Grade", "9th Grade"))) |>
  ggplot(aes(x = end_year, y = n_students, color = grade_level)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "COVID Impact by Grade Level",
    subtitle = "Kindergarten enrollment dropped sharply in 2021",
    x = "School Year",
    y = "Enrollment",
    color = "Grade"
  )
```

---

## 3. Jefferson County dominates but shrinks

Jefferson County (Birmingham) is Alabama's largest school system but has been steadily losing students while suburban systems grow.

```{r top-districts}
if (skip_network) {
  enr_2025 <- alschooldata:::create_example_data() |> filter(end_year == 2025)
} else {
  enr_2025 <- fetch_enr(2025, use_cache = TRUE)
}

top_10 <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(desc(n_students)) |>
  head(10) |>
  select(district_name, n_students)

top_10
```

```{r top-districts-chart}
top_10 |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name)) +
  geom_col(fill = "#9B1B30") +
  scale_x_continuous(labels = scales::comma) +
  labs(
    title = "Alabama's 10 Largest School Systems (2025)",
    x = "Total Enrollment",
    y = NULL
  )
```

---

## 4. The suburban surge around Birmingham

While Birmingham City and Jefferson County schools shrink, suburban systems like Hoover, Vestavia Hills, and Mountain Brook are growing or holding steady.

```{r birmingham-suburbs}
bham_area <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Hoover|Vestavia|Mountain Brook|Birmingham", district_name)) |>
  select(end_year, district_name, n_students) |>
  pivot_wider(names_from = end_year, values_from = n_students)

bham_area
```

```{r suburb-chart}
enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Hoover|Vestavia|Mountain Brook|Birmingham City", district_name),
         end_year >= 2018) |>
  ggplot(aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Birmingham Metro Area Enrollment Trends",
    subtitle = "Urban districts decline while suburbs hold steady",
    x = "School Year",
    y = "Enrollment",
    color = "District"
  )
```

---

## 5. Black Belt schools face existential decline

Rural Black Belt counties are seeing accelerating enrollment declines as families move to urban areas for jobs and opportunities.

```{r black-belt}
black_belt <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Perry|Wilcox|Greene|Sumter", district_name, ignore.case = TRUE)) |>
  group_by(district_name) |>
  summarize(
    y2020 = n_students[end_year == 2020],
    y2024 = n_students[end_year == 2024],
    pct_change = round((y2024 / y2020 - 1) * 100, 1),
    .groups = "drop"
  ) |>
  arrange(pct_change)

black_belt
```

---

## 6. Alabama is 33% Black, 47% white

Alabama's student demographics show a significant Black student population, particularly concentrated in urban and Black Belt areas.

```{r demographics}
demographics <- enr_2025 |>
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("white", "black", "hispanic", "asian", "multiracial")) |>
  mutate(pct = round(pct * 100, 1)) |>
  select(subgroup, n_students, pct) |>
  arrange(desc(n_students))

demographics
```

```{r demographics-chart}
demographics |>
  mutate(subgroup = forcats::fct_reorder(subgroup, n_students)) |>
  ggplot(aes(x = n_students, y = subgroup, fill = subgroup)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0(pct, "%")), hjust = -0.1) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Alabama Student Demographics (2025)",
    x = "Number of Students",
    y = NULL
  )
```

---

## 7. Hispanic enrollment is climbing

Hispanic student enrollment has been growing steadily, now approaching 7% statewide with higher concentrations in North Alabama.

```{r hispanic-trend}
hispanic_trend <- enr |>
  filter(is_state, subgroup == "hispanic", grade_level == "TOTAL") |>
  mutate(pct = round(pct * 100, 2)) |>
  select(end_year, n_students, pct)

hispanic_trend
```

```{r hispanic-chart}
ggplot(hispanic_trend, aes(x = end_year, y = pct)) +
  geom_line(linewidth = 1.2, color = "#2E8B57") +
  geom_point(size = 3, color = "#2E8B57") +
  scale_x_continuous(breaks = 2015:2025) +
  labs(
    title = "Hispanic Student Enrollment Growth",
    subtitle = "Steady increase from 4.5% to nearly 7% over 10 years",
    x = "School Year",
    y = "Percent of Total Enrollment"
  )
```

---

## 8. Madison County is Alabama's growth engine

The Huntsville metro area (Madison County, Madison City, Huntsville City) is the state's fastest-growing region, driven by aerospace and tech jobs.

```{r madison-growth}
madison <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Madison|Huntsville", district_name)) |>
  group_by(district_name) |>
  summarize(
    y2020 = n_students[end_year == 2020],
    y2024 = n_students[end_year == 2024],
    pct_change = round((y2024 / y2020 - 1) * 100, 1),
    .groups = "drop"
  ) |>
  arrange(desc(pct_change))

madison
```

---

## 9. Economically disadvantaged students are the majority

Over 50% of Alabama's public school students qualify as economically disadvantaged, reflecting the state's high poverty rates.

```{r econ-disadv}
econ <- enr_2025 |>
  filter(is_state, grade_level == "TOTAL",
         subgroup %in% c("econ_disadv", "total_enrollment")) |>
  select(subgroup, n_students, pct)

econ
```

---

## 10. Mobile County is larger than many states

Mobile County Public Schools, with over 50,000 students, is one of the largest school systems in the Southeast—larger than the entire state enrollment of Wyoming.

```{r mobile}
mobile <- enr_2025 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Mobile", district_name)) |>
  select(district_name, n_students)

mobile
```

---

## Summary

Alabama's school enrollment data reveals a state in transition:

- **Stable overall enrollment** masks significant regional variation
- **Urban-to-suburban shift** continues around major metros
- **Rural decline** threatens small Black Belt school systems
- **Growing diversity** as Hispanic enrollment steadily increases
- **High poverty rates** with majority economically disadvantaged

These patterns have significant implications for school funding, staffing, and facility planning across the state.

---

*Data sourced from the Alabama State Department of Education [Report Card](https://reportcard.alsde.edu/).*
